### 2015-07-23 Harris_2015/manuscript-materials/draft.Rmd
more

Estimating species interactions from observational data is one of the most
controversial tasks in community ecology. One difficulty is that a single
pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, two competing species would
ordinarily correlate negatively in space, but this effect can be reversed in
the presence of a third species that is capable of outcompeting both of them
when it is present. Here, I apply models from statistical physics, called
Markov networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix. Interactions
in these models can be estimated from observational data via maximum
likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and several existing approaches. The
Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions or
abiotic environmental effects largely overpowered them. A linear
approximation, based on partial covariances, also performed well as long as
the number of sampled locations exceeded the number of species in the data.
Indirect effects reliably caused a common null modeling approach to produce
incorrect inferences, however.

@@

While competition doesn't reliably leave the expected pattern in community
structure (a low co-occurrence rate at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
control is difficult to obtain from null models, which begin with the assumptions
that all the pairwise interactions are zero and thus don’t need to be controlled for.

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a framework
for understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks
have been used in many scientific fields for decades in similar contexts.
For example, a well-studied network named the Ising model has played an important
role in our understanding of physics [where nearby particles tend to align
magnetically with one another; @cipra_introduction_1987].
In spatial contexts, these models have been used to describe interactions between
adjacent grid cells [@harris_contact_1974; @gelfand_modelling_2005]. In
neurobiology, they have helped researchers identify neural connections from firing patterns
[@schneidman_weak_2006]. While ecologists explored some related approaches in the 1980's
[@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, this kind of approach is
worth a second look.

@@

\noindent  
***Conditional relationships and Markov networks.***
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The most familiar conditional relationship
is the partial correlation, which indicates the portion of the sample correlation
between two species that remains after controlling for other variables in the
data set [@albrecht_spatial_2001]. The example with the shrubs and trees in
Figure 1 shows how the two correlation measures can have opposite signs,
and suggests that the partial correlation is more relevant
for drawing inferences about species interactions (e.g. competition).
Markov networks extend this approach to non-Gaussian data, much as
generalized linear models do for linear regression [@lee_learning_2012].

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that
the presence of species $i$ contributes
to the log-probability of $\vec{y}$; it directly controls the prevalence of
species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
species $i$ and species $j$ contributes to the log-probability; it controls
the probability that the two species will be found together (Figure 2A, Figure 2B).
$\beta$ thus acts as an analog of the partial covariance, but for non-Gaussian
networks. Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce assemblages
that have many pairs of positively-associated species and relatively few pairs
of negatively-associated species (exactly as an ecologist might expect).

@@

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a normalizing constant,
$Z(\alpha, \beta)$, which is defined so that that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z(\alpha, \beta)$ exactly, as is done in this paper,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required spirals exponentially into the billions and beyond.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, e.g. via analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.

\noindent  
***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies. For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient as R code.  The rosalia package uses
standard convex optimization techniques to find the best values for $\alpha$ and $\beta$.

@@

I also simulated a second set of landscapes using a stochastic community
model based on generalized Lotka-Volterra dynamics, as described in the
Appendix. In these simulations, each species pair was randomly assigned to
either compete for a portion of the available carrying capacity (negative
interaction) or to act as mutualists (positive interaction).  In these
simulations, mutualism operated by mitigating the effects of intraspecific
competition on each partner's death rate.  For these analyses, I simulated
50 replicate landscapes, each with 25, 200, or 1600 sites and 20 species.

@@

For the Markov networks, I used the rosalia package [@harris_rosalia_2015], as described
above. Given the large number of parameters associated with some of the networks to
be estimated, I regularized the likelihood using a Cauchy prior distribution
[@gelman_weakly_2008] with a scale of 1 on the $\alpha$ and $\beta$ terms.

I also evaluated five alternative methods from the existing literature to calibrate the
performance of the Markov network as a method for estimating species interactions.
The first two alternative interaction measures were the sample correlations and the
partial correlations between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations are undefined
for landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@REF]. In the context of non-Gaussian
data, the partial correlation can be thought of as an approximation to the Markov network,
and is much easier to compute.

The third alternative, generalized linear models, can also be thought of as a computationally
efficient approximation to the Markov network [@REF].  Following @faisal_inferring_2010,
I fit regularized logistic regression models for each species, using the other species on the
landscape as predictors. [[Observational presence-absence data does not have enough degrees
of freedom to estimate this many parameters, so I symmetrized the pairwise correlations by
averaging the coefficient for predicting species $i$’s status from species $j$ and vice versa.]]

@@

Because the different methods mostly describe species interactions on different
scales (e.g. correlations versus $Z$ scores versus regression coefficients), I used
linear regression through the origin to rescale the different estimates produced
by each method so that they had a consistent interpretation. After rescaling each
method’s estimates, I calculated squared errors for each of the [[N]] species pairs
across all the simulated data sets and used these values to calculate the proportion of variance
explained for landscapes with 25, 200, or 1600 sampled locations (compared with a baseline
model that assumed all interaction strengths to be zero).

@@

\noindent  
***Larger landscapes.***
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering: overall, the Markov network
explained 54% of the "true" parameters' squared deviations from zero, followed
by partial covariances (33%), and sample covariances (22%). The null model
scores initially explained only 12%. After manually reducing the value of one
especially strong outlier ($Z=1004$, implying $p<10^{-1000000}$), this
increased to 17% (Appendix B).  Figure 3 reflects the adjusted version of the
results.

@@

\noindent  
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1). For cases where fitting a Markov network is computationally infeasible,
these results also indicate that partial covariances and generalized linear models (the two
methods that estimated conditional relationships rather than marginal ones) can both
provide useful approximations. The partial correlations' success on simulated
data may not carry over to real data sets, however; @loh_structure_2013 show that the linear
approximations can be less reliable in cases where the true interaction matrix
contains more structure (e.g. guilds or trophic levels).

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
null model must be interpreted with care [@roughgarden_competition_1983]. In
complex networks of ecological interactions (and even in small networks with
three species), it may simply not be possible to implicate individual species
pairs or specific ecological processes like competition by rejecting a
general-purpose null [@gotelli_empirical_2009]. Estimating pairwise
coefficients directly seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the
range of results shown in Figure 3, scientists in this field could easily
double their explanatory power by switching from null models to linear
regression and partial covariances, or triple it by switching to a Markov
network.

>>>>

Estimating species interactions from observational data is one of the most
controversial tasks in community ecology. One difficulty is that a single
pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, two competing species would
ordinarily correlate negatively in space, but this effect can be reversed in
the presence of a third species that is capable of outcompeting both of them
when it is present. Here, I apply models from statistical physics, called
Markov networks or Markov random fields, that can predict the direct and
indirect consequences of any possible species interaction matrix. Interactions
in these models can be estimated from observational data via maximum
likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and several existing approaches. The
Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions or
abiotic environmental effects largely overpowered them. Two computationally
efficient approximations, based on linear and generalized linear models, also performed
well. Indirect effects reliably caused a common null modeling approach and a Joint
Species Distribution Modeling approach to produce incorrect inferences, however.

@@

While competition doesn't reliably leave the expected pattern in community
structure (a low co-occurrence rate at the landscape level), it nevertheless does
leave a signal in the data (Figure 1C). Specifically, *among shaded sites*,
there will be a deficit of co-occurrences between the competing shrub species,
and *among unshaded sites*, there will also be such a deficit. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the assumptions
that all the pairwise interactions are zero and thus don’t need to be controlled for.
Nevertheless, null models have dominated this field for more than three decades
[@diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009].

In this paper, following recent work by @azaele_inferring_2010 and
@fort_statistical_2013, I show that Markov networks [undirected graphical models
also known as Markov random fields; @murphy_machine_2012] can provide a framework
for understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observational data. Markov networks
have been used in many scientific fields for decades in similar contexts, from physics
[where nearby particles interact magnetically with one another; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells interact with one another; @harris_contact_1974;
@gelfand_modelling_2005]. While community ecologists explored some related approaches in
the 1980's [@whittam_species_1981], computational limitations had previously imposed
severe approximations that produced unintelligible results [e.g.
"probabilities" greater than one; @gilpin_factors_1982].  Now that it is
computationally feasible to fit these models exactly, this kind of approach is
worth a second look.

@@

\noindent  
***Conditional relationships and Markov networks.***
Ecologists are often interested in learning about direct interactions between
species, controlling for the indirect influence of other species. In
statistical terms, this implies that ecologists want to estimate *conditional*
("all-else-equal") relationships, rather than *marginal* ("overall") relationships.
The example with the shrubs and trees in
Figure 1 shows how the two measures can have opposite signs,
and suggests that conditional relationships are more relevant
for drawing inferences about species interactions (e.g. competition).

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that
the presence of species $i$ contributes
to the log-probability of $\vec{y}$; it directly controls the prevalence of
species $i$. Similarly, $\beta_{ij}$ is the amount that the co-occurrence of
species $i$ and species $j$ contributes to the log-probability; it controls
the probability that the two species will be found together (Figure 2A, Figure 2B).
$\beta$ can be interpreted as an analog of the partial covariance, but for non-Gaussian
networks [@loh_structure_2013], or as similar to a coefficient in a logistic regression model
[@lee_learning_2012]. Because the relative probability of a presence-absence vector
increases when positively-associated species co-occur and decreases when
negatively-associated species co-occur, the model tends to produce assemblages
that have many pairs of positively-associated species and relatively few pairs
of negatively-associated species (exactly as an ecologist might expect).

@@

Estimating the marginal relationships predicted by a Markov network is more
difficult than estimating conditional relationships, because doing so requires
absolute probability estimates. Turning the relative probability given by
Equation 1 into an absolute probability entails scaling by a normalizing constant,
$Z$, which is defined so that that the probabilities of all
possible assemblages that could be produced by the model sum to one (bottom of
Figure 2B). Calculating $Z$ exactly, as is done in these analyses,
quickly becomes infeasible as the number of species increases: with
$2^N$ possible assemblages of $N$ species, the number of bookkeeping operations
required spirals exponentially into the billions and beyond.
Numerous techniques are available for working with Markov networks that keep
the computations tractable, e.g. via analytic approximations
[@lee_learning_2012] or Monte Carlo sampling [@salakhutdinov_learning_2008],
but they are beyond the scope of this paper.  Alternatively, as noted below, some
common linear and generalized linear methods can be used as computationally efficient
approximations to the full Markov network.

\noindent  
***Inferring $\alpha$ and $\beta$ coefficients from presence-absence data.***
In the previous two sections, the values of $\alpha$ and $\beta$ were known. In practice,
however, ecologists will often need to estimate these parameters from co-occurrence data.
When the number of species is reasonably small, one can compute exact maximum
likelihood estimates for all of the $\alpha$ and $\beta$ coefficients by optimizing
$p(\vec{y}; \alpha, \beta)$. Fully-observed Markov
networks like the ones considered here have unimodal likelihood surfaces
[@murphy_machine_2012], ensuring that this procedure will always converge on
the global maximum. This maximum is the unique combination of $\alpha$ and $\beta$
coefficients that would be expected to produce exactly the observed co-occurrence
frequencies (i.e. it matches the sufficient statistics of the data to the sufficient statistics of the
model distribution). For the analyses in this paper, I used the *rosalia* package
[@harris_rosalia_2015] for the R programming language [@r_core_team_r_2015] to
define the objective function and gradient as R code.  The rosalia package uses
standard convex optimization techniques to find the best values for $\alpha$ and $\beta$.

@@

I also simulated a second set of landscapes using a stochastic community
model based on generalized Lotka-Volterra dynamics, as described in the
Appendix. In these simulations, each species pair was randomly assigned to
either compete for a portion of the available carrying capacity (negative
interaction) or to act as mutualists (positive interaction).  In these
simulations, mutualism operated by mitigating the effects of intraspecific
competition on each partner's death rate.  For these analyses, I simulated
landscapes with up to 20 species and 25, 200, or 1600 sites (50 replicates per
landscape size; see Appendix).

@@

For the Markov networks, I used the rosalia package [@harris_rosalia_2015], as described
above. Given the large number of parameters associated with some of the networks to
be estimated, I regularized the likelihood using a logistic prior distribution
with a scale of 1 on the $\alpha$ and $\beta$ terms.  This convex prior distribution has a
similar shape to the Laplace distribution used in LASSO regularization (especially in the tails),
but does not force the estimates to be exactly zero.

I also evaluated five alternative methods from the existing literature to calibrate the
performance of the Markov network as a method for estimating species interactions.
The first two alternative interaction measures were the sample correlations and the
partial correlations between each pair of species' data vectors on the landscape
[@albrecht_spatial_2001; @faisal_inferring_2010]. Because partial correlations are undefined
for landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@REF]. In the context of non-Gaussian
data, the partial correlation can be thought of as a computationally efficient approximation to the
full Markov network model [@loh_structure_2013].

The third alternative, generalized linear models, can also be thought of as a computationally
efficient approximation to the Markov network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized logistic regression models for each species, using
the other species on the landscape as predictors. [[Observational presence-absence data does
not have enough degrees of freedom to estimate this many parameters, so I symmetrized the
pairwise correlations by averaging the coefficient for predicting species $i$’s status from
species $j$ and vice versa.]]

@@

Because the different methods mostly describe species interactions on different
scales (e.g. correlations versus $Z$ scores versus regression coefficients), I used
linear regression through the origin to rescale the different estimates produced
by each method so that they had a consistent interpretation. After rescaling each
method’s estimates, I calculated squared errors between the scaled interaction estimates and
“true” interaction values for each of the [[N]] species pairs
across all the simulated data sets. These squared errors determined the proportion of variance
explained for different combinations of method and landscape size (compared with a baseline
model that assumed all interaction strengths to be zero).

@@

\noindent  
***Twenty species.***
The accuracy of the four evaluated methods varied substantially, depending on
the parameters that produced the simulated communities (Figure 3). In general,
however, there was a consistent ordering…

The three methods that estimated marginal relationships among species and did
not explicitly control for indirect species interactions (the null model, the joint species
distribution model, and the sample correlation coefficients) differed from one another
in their $R^2$ values, but their estimates were surprisingly similar to one another.
In particular, the correlation coefficients were extremely similar to the estimates
from the other two methods (Spearman’s rho = X for the joint species distribution
model and Y for the null model).

@@

\noindent  
The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by indirect
effects (Figure 1). For cases where fitting a Markov network is computationally infeasible,
these results also indicate that partial covariances and generalized linear models (the two
methods that estimated conditional relationships rather than marginal ones) can both
provide useful approximations. The partial correlations' success on simulated
data may not carry over to real data sets, however; @loh_structure_2013 show that the linear
approximations can be less reliable in cases where the true interaction matrix
contains more structure (e.g. guilds or trophic levels). Similarly, the approximation involved
in using separate generalized linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes [@lee_learning_2012].

@@

Finally, the results presented here have important implications for
ecologists' continued use of null models to draw inferences about species
interactions. Null and neutral models can be very useful for clarifying our
thinking about the numerical consequences of species' richness and abundance
patterns [@harris_occupancy_2011; @xiao_strong_2015], but deviations from a
null model must be interpreted with care [@roughgarden_competition_1983]. In
complex networks of ecological interactions (and even in small networks with
three species), it may simply not be possible to implicate individual species
pairs or specific ecological processes like competition by rejecting a
general-purpose null [@gotelli_empirical_2009]. Estimating pairwise
coefficients directly seems like a much more promising approach: to the extent
that the models' relative performance on real data sets is similar to the
range of results shown in Figure 3, scientists in this field could easily
double their explanatory power by switching from null models to partial correlations or
generalized linear models, or triple it by switching to a Markov network. Regardless
of the specific methods ecologists ultimately choose, the most important consideration
is clearly that we need to control for indirect effects and estimate conditional relationships
between species, rather than marginal ones.
